Constraints on Cosmic-Ray Particle Populations, Magnetic Fields, and Dark Matter from VHE Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS

### 2011-09-09 ms.tex
Minor changes

Moreover, a substantial value of $X_\CR$ can bias hydrostatic cluster masses as well as the
temperature decrement of the cosmic microwave background (CMB) in the direction of a galaxy cluster
due to the Sunyaev-Zel'dovich effect which is proportional to the line-of-sight integral of the
thermal electron pressure. This could then severely jeopardize the use of clusters to determine
cosmological parameters. Comparing X-ray and optical potential profiles in the centers of galaxy
clusters yields an upper limit of 20--30 \% of non-thermal pressure (that can be composed of CRs,
magnetic fields or turbulence) relative to the thermal gas pressure
\citep{article:Churazov_etal:2008, article:Churazov_etal:2010}. An analysis that compares spatially
resolved weak gravitational lensing and hydrostatic X-ray masses for a sample of 18 galaxy clusters
detects a deficit of the hydrostatic in comparison to the lensing mass of $20 \%$ at $R_{500}$ --
within which the mean density is 500 times the critical density -- suggesting again a substantial
non-thermal pressure contribution on large scales \citep{article:Mahdavi_etal:2008}. High-energy
gamma-ray emission is a complementary method of constraining the pressure contribution of CRs that
is most sensitive to the cluster core region. However, it assumes that the CR component is fully
mixed with the ICM and almost insensitive in detecting a two-phase structure of CRs and the thermal
ICM.  A fraction $X_\CR$ of only a few percent is required in order to produce a gamma-ray flux
observable with current generation gamma-ray telescopes rendering this technique at least as
sensitive as the dynamical and hydrostatic method (which are however more general in that they are
sensitive to any non-thermal pressure component).

@@

The analysis of this work follows the same procedure as described in detail in
\citet{article:Abdo_etal:2009}. To eliminate albedo gamma rays from the Earth limb, a zenith angle
cut of 105$^{\circ}$ is applied. The \emph{Fermi} Science Tools version v9r23p1 and Instrumental
Response Functions (IRFs) P6\_V11\_Diffuse (a model of the spatial distribution of photon events
calibrated  in flight) were used throughout this work. Only photons above 1 GeV are considered,
where the constraints on the flux are most meaningful. The \emph{Fermi}-LAT collaboration estimates
the systematic uncertainties at 10 GeV to be around 20\%.
\footnote{\url{http://fermi.gsfc.nasa.gov/ssc/data/analysis/LAT\_caveats.html}} Flux upper limits
at the 99\% confidence level in the energy range 1--30 GeV, assuming both a unresolved, point-like
or spatially extended emission, are tabulated in Table \ref{table:fermi}. The gamma-ray emission is
assumed to follow a power-law distribution with a photon index $\alpha=2.1$, 2.3, and $2.5$.

@@

We adopt the electron density profile for the Coma cluster that has been inferred from ROSAT X-ray
observations \citep{article:BrielHenryBohringer:1992} and model the spatial and spectral
distribution of CRs according to the model by \citet{article:PinzkePfrommer:2010} that neglects the
contribution of cluster galaxies. Figure \ref{fig:spectrum} shows the expected spectrum of Coma,
normalized to the total flux within the virial radius (dotted lines) and the angular apertures
tested in this study (solid lines). The gamma-ray flux limit of {\em Fermi} in the energy interval
of 1--3 GeV ($<0.4^\circ$) is most constraining. However, since the model flux within the finite
energy range of 1--3 GeV is smaller by a factor of 1.5 in comparison to the total flux above 1 GeV,
it falls below the flux upper limits which do not yet constrain the simulation model. In our
further analysis, we use the most constraining {\em Fermi} flux limits in the energy interval of
1-3 GeV as well as the gamma-ray flux limits of VERITAS in the energy range above 220 GeV.

@@

Table \ref{table:constraints} shows the expected gamma-ray fluxes within the regions of interest.
In the case of VERITAS, for the most constraining regions tested (within an aperture of radius
0.2$^{\circ}$), those are a factor of 7.2 below the inferred upper limits of VERITAS (see Table
\ref{table:results} and assuming a spectral index of $\alpha=2.3$ which matches the simulated one
at energies $E_\gamma=200$ GeV). Hence we scale the averaged relative CR pressure of our model by
that factor, keep the spatial behavior and obtain an integrated limit of the relative CR pressure
of $X_{\CR}<0.105$ within 0.2$^{\circ}$ that translates to a limit within the cluster virial radius
of $X_\CR<0.152$ (solid lines of Figure \ref{fig:xcr}). Equivalently, the most constraining
{\em Fermi} upper limit in the energy interval of 1--3 GeV ($<0.4^\circ$) is a factor of 1.1 larger
than our model predictions (assuming $\alpha=2.1$ which is very close to the simulated spectral
index for the energy range of 1-3 GeV). Scaling our integrated relative CR pressure profile yields
the constraint of $X_{\CR}<0.018$ within 0.4$^{\circ}$ that translates to a limit within the
cluster virial radius of $X_\CR<0.024$ (dashed lines of Figure \ref{fig:xcr}).

@@

The expected gamma-ray flux due to self-annihilations of WIMPs is given by 
\begin{equation}
\frac{d\Phi_{\gamma}(\Delta\Omega,E)}{dE}=
\frac{\expval{\sigma v}}{8\pi m_{\chi}^{2}}\frac{dN(E,m_{\chi})}{dE} J(\Delta\Omega),
\label{eqn:WIMPflux}
\end{equation}
where $\expval{\sigma v}$ is the thermally averaged product of the total self-annihilation cross
section and the WIMP velocity, $m_{\chi}$ is the WIMP mass, $dN(E,m_{\chi})/dE$ is the differential
gamma-ray yield per annihilation, $\Delta\Omega$ is the observed solid angle, and $J$ is the
astrophysical factor - a dimensionless factor which determines the DM annihilation rate and depends
on the DM distribution.

@@

In this work we have adopted the DM density profile by Navarro-Frenk-White
\citep[NFW;][]{article:NavarroFrenkWhite:1997},
\begin{equation}
\rho_{\chi}(r)=\rho_{s}\left(\frac{r}{r_{s}}\right)^{-1}\left(1+\frac{r}{r_{s}}\right)^{-2},
\end{equation} 
which parametrizes the distribution of DM by the scale radius $r_{s}$ and the scale density
$\rho_{s}$. [how where these calculated?]

@@

The {\em Fermi} flux limits constrain $X_\CR$ to be below 0.017--0.025 (for $\alpha$ varying
between 2.3 and 2.1), only weakly depending on the assumed CR spectral index. Assuming $\alpha=2.1$
which is very close to the simulated spectral index for the energy range of 1--3 GeV, we obtain an
almost coincident constraint with our simulation-based model within the virial radius of
$X_\CR<0.024$. That constraint improves to $X_{\CR}<0.018$ for an aperture of 0.4$^\circ$
corresponding to a physical scale of $R \simeq R_{200}/3 \simeq 660$~kpc. These are encouraging
results in that we constrain the CR pressure (of a phase that is fully mixed with the ICM) to be at
most a small fraction of the overall pressure, $<0.025$. As a result, hydrostatic cluster masses
and the total Comptonization parameter due to the Sunyaev-Zel'dovich effect attain at most a very
small bias due to CRs.

@@

\begin{figure*}
\begin{center}
\scalebox{1.0}{\plotone{f4.eps}}
\end{center}
\caption{Integral flux upper limits (this work, Tables~\ref{table:results} and \ref{table:fermi})
for different integration radii (shown with different colors), assuming $\alpha=2.1$. These are
compared to simulated integrated spectra of VHE gamma-ray emission, $E_\gamma F_\gamma(>E_\gamma)$,
from the Coma cluster (again within different apertures). Note that the {\em Fermi} limit for the
energy interval of 1--3 GeV within the aperture of 0.4$^\circ$ is most constraining. Since the
model flux within the finite energy range of 1--3 GeV is smaller by a factor of 1.5 in comparison
to the total flux above 1 GeV, it falls below the flux upper limits which do not yet constrain the
simulation model.}
\label{fig:spectrum}
\end{figure*}

>>>>

Moreover, a substantial value of $X_\CR$ can bias hydrostatic cluster masses as well as the
temperature decrement of the cosmic microwave background (CMB) in the direction of a galaxy cluster
due to the Sunyaev-Zel'dovich effect which is proportional to the line-of-sight integral of the
thermal electron pressure. This could then severely jeopardize the use of clusters to determine
cosmological parameters. Comparing X-ray and optical potential profiles in the centers of galaxy
clusters yields an upper limit of 20-30 \% of non-thermal pressure (that can be composed of CRs,
magnetic fields or turbulence) relative to the thermal gas pressure
\citep{article:Churazov_etal:2008, article:Churazov_etal:2010}. An analysis that compares spatially
resolved weak gravitational lensing and hydrostatic X-ray masses for a sample of 18 galaxy clusters
detects a deficit of the hydrostatic in comparison to the lensing mass of $20 \%$ at $R_{500}$ --
within which the mean density is 500 times the critical density -- suggesting again a substantial
non-thermal pressure contribution on large scales \citep{article:Mahdavi_etal:2008}. High-energy
gamma-ray emission is a complementary method of constraining the pressure contribution of CRs that
is most sensitive to the cluster core region. However, it assumes that the CR component is fully
mixed with the ICM and almost insensitive in detecting a two-phase structure of CRs and the thermal
ICM.  A fraction $X_\CR$ of only a few percent is required in order to produce a gamma-ray flux
observable with current generation gamma-ray telescopes rendering this technique at least as
sensitive as the dynamical and hydrostatic method (which are however more general in that they are
sensitive to any non-thermal pressure component).

@@

The analysis of this work follows the same procedure as described in detail in
\citet{article:Abdo_etal:2009}. To eliminate albedo gamma rays from the Earth limb, a zenith angle
cut of 105$^{\circ}$ is applied. The \emph{Fermi} Science Tools version v9r23p1 and Instrumental
Response Functions (IRFs) P6\_V11\_Diffuse (a model of the spatial distribution of photon events
calibrated  in flight) were used throughout this work. Only photons above 1 GeV are considered,
where the constraints on the flux are most meaningful. The \emph{Fermi}-LAT collaboration estimates
the systematic uncertainties at 10 GeV to be around 20\%.
\footnote{\url{http://fermi.gsfc.nasa.gov/ssc/data/analysis/LAT\_caveats.html}} Flux upper limits
at the 99\% confidence level in the energy range 1-30 GeV, assuming both a unresolved, point-like
or spatially extended emission, are tabulated in Table \ref{table:fermi}. The gamma-ray emission is
assumed to follow a power-law distribution with a photon index $\alpha=2.1$, 2.3, and $2.5$.

@@

We adopt the electron density profile for the Coma cluster that has been inferred from ROSAT X-ray
observations \citep{article:BrielHenryBohringer:1992} and model the spatial and spectral
distribution of CRs according to the model by \citet{article:PinzkePfrommer:2010} that neglects the
contribution of cluster galaxies. Figure \ref{fig:spectrum} shows the expected spectrum of Coma,
normalized to the total flux within the virial radius (dotted lines) and the angular apertures
tested in this study (solid lines). The gamma-ray flux limit of {\em Fermi} in the energy interval
of 1-3 GeV ($<0.4^\circ$) is most constraining. However, since the model flux within the finite
energy range of 1-3 GeV is smaller by a factor of 1.5 in comparison to the total flux above 1 GeV,
it falls below the flux upper limits which do not yet constrain the simulation model. In our
further analysis, we use the most constraining {\em Fermi} flux limits in the energy interval of
1-3 GeV as well as the gamma-ray flux limits of VERITAS in the energy range above 220 GeV.

@@

Table \ref{table:constraints} shows the expected gamma-ray fluxes within the regions of interest.
In the case of VERITAS, for the most constraining regions tested (within an aperture of radius
0.2$^{\circ}$), those are a factor of 7.2 below the inferred upper limits of VERITAS (see Table
\ref{table:results} and assuming a spectral index of $\alpha=2.3$ which matches the simulated one
at energies $E_\gamma=200$ GeV). Hence we scale the averaged relative CR pressure of our model by
that factor, keep the spatial behavior and obtain an integrated limit of the relative CR pressure
of $X_{\CR}<0.105$ within 0.2$^{\circ}$ that translates to a limit within the cluster virial radius
of $X_\CR<0.152$ (solid lines of Figure \ref{fig:xcr}). Equivalently, the most constraining
{\em Fermi} upper limit in the energy interval of 1-3 GeV ($<0.4^\circ$) is a factor of 1.1 larger
than our model predictions (assuming $\alpha=2.1$ which is very close to the simulated spectral
index for the energy range of 1-3 GeV). Scaling our integrated relative CR pressure profile yields
the constraint of $X_{\CR}<0.018$ within 0.4$^{\circ}$ that translates to a limit within the
cluster virial radius of $X_\CR<0.024$ (dashed lines of Figure \ref{fig:xcr}).

@@

The expected gamma-ray flux due to self-annihilations of WIMPs is given by 
\begin{equation}
\frac{d\Phi_{\gamma}(\Delta\Omega,E)}{dE}=
\frac{\expval{\sigma v}}{8\pi m_{\chi}^{2}}\frac{dN(E,m_{\chi})}{dE} J(\Delta\Omega),
\label{eqn:WIMPflux}
\end{equation}
where $\expval{\sigma v}$ is the thermally averaged product of the total self-annihilation cross
section and the WIMP velocity, $m_{\chi}$ is the WIMP mass, $dN(E,m_{\chi})/dE$ is the differential
gamma-ray yield per annihilation, $\Delta\Omega$ is the observed solid angle, and $J$ is a
dimensionless factor, referred to as the astrophysical factor, which determines the DM annihilation
rate and depends on the DM distribution.

@@

In this work we have adopted the DM density profile by Navarro-Frenk-White
\citep[NFW;][]{article:NavarroFrenkWhite:1997},
\begin{equation}
\rho_{\chi}(r)=\rho_{s}\left(\frac{r}{r_{s}}\right)^{-1}\left(1+\frac{r}{r_{s}}\right)^{-2},
\end{equation} 
which parametrizes the distribution of DM by the scale radius $r_{s}$ and the scale density
$\rho_{s}$.

@@

The {\em Fermi} flux limits constrain $X_\CR$ to be below 0.017-0.025 (for $\alpha$ varying
between 2.3 and 2.1), only weakly depending on the assumed CR spectral index. Assuming $\alpha=2.1$
which is very close to the simulated spectral index for the energy range of 1-3 GeV, we obtain an
almost coincident constraint with our simulation-based model within the virial radius of
$X_\CR<0.024$. That constraint improves to $X_{\CR}<0.018$ for an aperture of 0.4$^\circ$
corresponding to a physical scale of $R \simeq R_{200}/3 \simeq 660$~kpc. These are encouraging
results in that we constrain the CR pressure (of a phase that is fully mixed with the ICM) to be at
most a small fraction of the overall pressure, $<0.025$. As a result, hydrostatic cluster masses
and the total Comptonization parameter due to the Sunyaev-Zel'dovich effect attain at most a very
small bias due to CRs.

@@

\begin{figure*}
\begin{center}
\scalebox{1.0}{\plotone{f4.eps}}
\end{center}
\caption{Integral flux upper limits (this work, Tables~\ref{table:results} and \ref{table:fermi})
for different integration radii (shown with different colors), assuming $\alpha=2.1$. These are
compared to simulated integrated spectra of VHE gamma-ray emission, $E_\gamma F_\gamma(>E_\gamma)$,
from the Coma cluster (again within different apertures). Note that the {\em Fermi} limit for the
energy interval of 1-3 GeV within the aperture of 0.4$^\circ$ is most constraining. Since the
model flux within the finite energy range of 1-3 GeV is smaller by a factor of 1.5 in comparison
to the total flux above 1 GeV, it falls below the flux upper limits which do not yet constrain the
simulation model.}
\label{fig:spectrum}
\end{figure*}
